---
title: "A cursory view on Generalised Additive (Mixed) Models (GAMMs)"
author: "Michele Gubian"
date: "9/17/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(mgcv)
library(itsadug)

```

# From coefficients ($\beta$'s) to shapes

## A naive approach

Suppose we have several realisations of a pitch accent in two conditions.
For simplicity:

* no random factors, e.g. all utterances from the same speaker
* all utterances examined on the same time interval, 0-2 sec
* or we might imagine that limear time normalisation was applied


```{r echo=FALSE, message=FALSE}
home_dir <- "/homes/m.gubian/Dokumente/Courses/Graz2020/FPCA-phonetics-workshop"
data_dir <- file.path(home_dir,'data')
ex <- 1 
curves <- read_csv(file.path(data_dir, paste("ex1D", ex, "csv", sep = '.')))
nCurves <- curves %>% select(curveId) %>% n_distinct()
curves %<>% mutate(Category = factor(Category))


# plot a few curves
ggplot(curves %>% filter(curveId %in% sample(nCurves, 20))) +
  aes(x = time, y = y, group = curveId, color = Category) +
  geom_line() +
  theme_light() +
  theme(text = element_text(size = 15))

```
We could try something like this and use lm():

$$
y_i = \beta_0 + \beta_{0, P} + \beta_1 t_i + \beta_{1, P} t_i +   \beta_2 t_i^2 + \beta_{2, P} t_i^2 + 
\beta_3 t_i^3 + \beta_{3, P} t_i^3 + ... + \epsilon_i 
$$
where:

* $P$ indicates the PEAK category, while the default (control) level is NO_PEAK
* instead of $x$ we have $t$ for time
* $i$ is NOT an index for an entire curve, but for a single sample

(You can try it at home)

### Problems

* We don't know where to stop with power terms
  - but we could use model selection 
* Pure power series are really bad at interpolating general shapes
  - because Taylor's theorem is about approximating curves near a specific point, not globally
* How are we going to use the $\beta$ terms for insight and interpretation?

### Solutions adopted by GAMs

* Use **SPLINES** as base (polynomial) functions, because they are good for interpolating any shape
* Use **regularisation** and **Cross-Validation** to determine complexity (how many functions) and smoothness 
* **Interpret shapes**, not coefficients

[Slides by Simon Wood](https://www.maths.ed.ac.uk/~swood34/talks/Beijing15.pdf)

The model looks like:

$$
y_i = f(t_i) + f_P(t_i) + \epsilon_i 
$$
where each $f()$ is a SPLINE:
$$
f(t) = \sum_{k=1}^K \beta_k b_k(t)
$$
In the GAMs lingo, the $f()$ are called **smooths**.

### GAMs in R

Package **mgcv**.

```{r}
gam1 <- bam(y ~ Category + s(time, by = Category), data = curves)
summary(gam1)

plot_smooth(gam1, view = "time", plot_all = "Category")
plot_diff(gam1, view = "time", comp=list(Category = c("PEAK", "NO_PEAK")))
```

### Random smooths

Just like we moved from lm() to lmer(), we can build mixed-effects GAMs, or GAMMs.

* Random effects become random smooths
* These can model 'intercept' and/or 'slopes', i.e.:
  - intercept: e.g. each speaker has a specific f0 shape that departs from the mean f0 contour for all speakers
  - slope: e.g. each speaker has a specific way of realising a contrast expressed in the entire f0 contour
* Smooth (functional) correlation between smooth random intercepts and slopes cannot be modelled with the mgcv package


### What is special about time-varying signals?

* What is the difference between time-series and other types of functional data?
* When we provide information (to the model) about the grouping of smooths induced by having different subjects, is there any other relationship in the data we would like to represent?

We would like to indicate which $(t_i, y_i)$ samples belong to the same curve! There are two ways of doing it:

* Define another (nested) level of random smooth, call it the token
  - but it is rarely done in practice, as it is likely to fail to converge (but see [this paper by van Rij et al.](https://journals.sagepub.com/doi/pdf/10.1177/2331216519832483))
* Alternatively, enrich $\epsilon$ with an AR1 model to express correlation of adjacent samples
  - $\epsilon_i = \rho \epsilon_{i-1} + \psi_i$

### Practical advice

* Convert all character variables to factor
* Convert binary variables to integer
* GAMMs with random smooth slopes are **very very slow**, use a server, not your
  laptop, and use the nthreads option of bam()

  
  

### Resources

* Tutorial papers: [Martijn Wieling](http://martijnwieling.nl/files/GAM-tutorial-Wieling.pdf), [Marton Soskuthy](http://eprints.whiterose.ac.uk/113858/2/1703_05339v1.pdf)
* Online tutorials: [Joseph Roy](https://jroy042.github.io/nonlinear/), [Jacolien van Rij](http://jacolienvanrij.com/Tutorials/GAMM.html), [Peter Laurinec](https://petolau.github.io/Analyzing-double-seasonal-time-series-with-GAM-in-R/), [Noam Ross](https://noamross.github.io/gams-in-r-course/), [Michael Clark](https://m-clark.github.io/generalized-additive-models/)
