---
title: "Notes on ex1D"
author: "Michele Gubian"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load relevant libraries and set up.
```{r}
library(fda)
library(funData)
library(MFPCA)
library(tidyverse)
library(landmarkregUtils)

mytheme <- theme_light() +
  theme(text = element_text(size = 16))
Category.colors <- c(A = "darkslategray", B = "orangered")
data_dir <- "data"

```
Load an example (`ex`), apply linear time normalization.
```{r}
ex <- 1 # change according to ex number
raw_curves <- readRDS(file.path(data_dir, str_c("ex1D", ex, "rds", sep = '.'))) %>% ungroup() %>% 
  mutate(across(c(curveId, Category), ~ factor(.x)))

# Linear time normalization:
# resample all curves on the same regular grid, 
# grid is conventionally between 0 and 1.
nSamples <- 40
grid <- seq(0, 1, length.out = nSamples)
curves <- raw_curves %>% 
  # group by all the factors at the level of curveId or higher (e.g. speaker)
  group_by(curveId, Category) %>% 
  # shift all curves to start from 0
  mutate(time = time - min(time)) %>%
  reframe(tibble(
    y = approx(time, y, n = nSamples)$y,
    time = grid
  ))
```
Plot a few curves
```{r}
# plot a few curves
subset_curveId <- raw_curves %>%
  ungroup() %>% 
  distinct(curveId) %>%
  slice_sample(n = 20)

ggplot(curves %>% inner_join(subset_curveId, by = "curveId")) +
  aes(x = time, y = y, group = curveId, color = Category) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values=Category.colors) +
  mytheme  +
  theme(legend.position = "bottom")

```


Compute FPCA
```{r}
# build a funData object
curvesFun <- long2irregFunData(curves, id = "curveId", time = "time", value = "y") %>% 
    as.funData()

# Compute FPCA
fpca <- PACE(curvesFun)

```
Plot PC curves
```{r}
# scores st. dev.
sdFun <- fpca$values %>% sqrt()
# PC curves to be plotted
nPC <- 2
PCcurves <- expand_grid(PC = 1:nPC,
                        fractionOfStDev = seq(-1, 1, by=.25)) %>%
  group_by(PC, fractionOfStDev) %>%
  reframe(
    funData2long1(fpca$mu + fractionOfStDev * sdFun[PC] * fpca$functions[PC],
                  time = "time", value = "y")
  )
# Plot
ggplot(PCcurves) +
  aes(x = time, y = y, group = fractionOfStDev, color = fractionOfStDev) +
  geom_line() +
  scale_color_gradient2(low = "blue", mid = "grey", high = "orangered",
                        breaks = c(-1, 0 , 1)) +
  facet_wrap(~ PC, nrow = 1,
             labeller = labeller(PC = ~ str_glue("PC{.x}"))) +
  labs(color = expression(frac(s[k], sigma[k]))) +
  xlab("registered time") +
  geom_line(data = PCcurves %>% filter(fractionOfStDev == 0), color = 'black', linewidth = 1.2) +
  mytheme +
  theme(legend.position = "bottom")

```

Plot the mean curve
```{r}
# quick way (not ggplot)
fpca$mu %>% plot()
# nicer way
fpca$mu %>% funData2long1(time = "time", value = "y") %>% ggplot() +
  aes(x = time, y = y) +
  geom_line()
```

Plot PC1
```{r}
fpca$functions[1] %>% funData2long1(time = "time", value = "y") %>% ggplot() +
  aes(x = time, y = y) +
  geom_line()
```

Print the PC scores of the first few curves
```{r}
fpca$scores %>% head()
```
Reconstruct the first curve (first row of scores matrix) up to the first two PC scores (first two columns of scores matrix). Compare to original curve.
```{r}
orig_curve <- curvesFun[1] %>% funData2long1(value = "y")
rec_curve <- (fpca$mu + fpca$scores[1,1] * fpca$functions[1] + 
                fpca$scores[1,2] * fpca$functions[2]) %>% funData2long1(value = "y")
# join the curves into one long format tibble and plot
bind_rows(orig_curve %>% mutate(curve = "original"),
          rec_curve %>% mutate(curve = "reconstructed")) %>% 
  ggplot(aes(time, y, color = curve)) +
  geom_line()
```

Plot PC scores scatterplot
```{r}
PCscores <- fpca$scores %>%
  `colnames<-`( paste0("s", 1:fpca$npc)) %>%
  as_tibble() %>%
  bind_cols(curves %>% distinct(curveId, Category), .)

# scatterplot PC scores s1 and s2 by Category
ggplot(PCscores) +
  aes(x = s1, y = s2, color = Category) +
  geom_point() +
  scale_color_manual(values=Category.colors) +
  mytheme +
  theme(legend.position = "right")
```

Pick two random curves from the dataset, call them  `curve1` and `curve2`. Plot the curves. Then mark the corresponding points on the scatterplot.
```{r}
id1 <- 1
id2 <- 55
curve1 <- curves %>% filter(curveId==id1)
curve2 <- curves %>% filter(curveId==id2)
ggplot(mapping = aes(time, y)) +
  geom_line(data = curve1, color = "magenta") +
  geom_line(data = curve2, color = "chartreuse") 

ggplot(PCscores) +
  aes(x = s1, y = s2) +
  geom_point() +
  geom_point(data = PCscores %>% filter(curveId==id1), color = "magenta", size = 3, shape = 18) +
  geom_point(data = PCscores %>% filter(curveId==id2), color = "chartreuse", size = 3, shape = 18) +
  mytheme +
  theme(legend.position = "right")
```

Construct a new curve based on  `curve1` and `curve2`, call it `curveX`. The new curve is obtained by averaging the PC scores of  `curve1` and `curve2`. Consider the first two PC scores. Plot `curveX`'s PC scores on the scatterplot. Plot `curveX` together with  `curve1` and `curve2`.
```{r}
scoresX <- PCscores %>% filter(curveId %in% c(id1, id2)) %>% summarise(across(c(s1, s2), mean))
ggplot(PCscores) +
  aes(x = s1, y = s2) +
  geom_point() +
  geom_point(data = PCscores %>% filter(curveId==id1), color = "magenta", size = 3, shape = 18) +
  geom_point(data = PCscores %>% filter(curveId==id2), color = "chartreuse", size = 3, shape = 18) +
  geom_point(data = scoresX, color = "blue", size = 3, shape = 18) +
  geom_segment(x = PCscores %>% filter(curveId==id1) %>% pull(s1),
               xend = PCscores %>% filter(curveId==id2) %>% pull(s1),
               y = PCscores %>% filter(curveId==id1) %>% pull(s2),
               yend = PCscores %>% filter(curveId==id2) %>% pull(s2),
               linetype = "dashed") +
  mytheme +
  theme(legend.position = "right")

curveX <- (
  fpca$mu + (scoresX %>% pull(s1)) * fpca$functions[1] +
    (scoresX %>% pull(s2)) * fpca$functions[2]
) %>% funData2long1(value = "y")
ggplot(mapping = aes(time, y)) +
  geom_line(data = curve1, color = "magenta") +
  geom_line(data = curve2, color = "chartreuse") +
  geom_line(data = curveX, color = "blue")
```

